\documentclass{article}
\usepackage{geometry}
\usepackage{graphicx} % Required for inserting images
\usepackage{amsmath, amsthm, amssymb}
\usepackage{mathtools}
\DeclarePairedDelimiter\ceil{\lceil}{\rceil}
\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}
\usepackage{parskip}
\newgeometry{vmargin={15mm}, hmargin={24mm,34mm}}
\theoremstyle{definition} 
\newtheorem{definition}{Definition}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{proposition}[theorem]{Proposition}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\xtoc}{\lvert x \rvert ^{c}}
\newcommand{\dprime}{\prime\prime}
\newcommand{\opt}{\text{opt}}
\DeclareMathOperator*{\E}{\mathbb{E}}

\title{Schwarz-Zippel Lemma}
\date{March 2024}
\author{Boran Erol}

\begin{document}

\maketitle

Starting from Lecture 10 of Ryan O'Donnell CS Theory Toolkit. He also has some resources he recommends.

We can construct and work with $F_{q}$ in $polylog(q)$ time. In certain rare cases,
we need randomization.

For $p$ a prime, $F_{p}$ is a field. In particular, it's a Euclidean domain, so you
can efficiently find the inverse using the extended GCD algorithm by solving 
Bezout's Identity.

An important open problem is TCS is whether we can compute the gcd of two numbers
using an efficient parallel algorithm.

One catch to working with $F_{p}$ is that we sometimes need to find some prime $p$ that's
of exponential size in $n$. We can work with such fields since representing elements of this field
is polynomial time. However, how do we find such a $p$ in the first place? The answer is disappointing:
sample a random number and test whether it's prime. We don't know whether a deterministic
algorithm exists for finding primes. Notice that this is a Las Vegas algorithm and it's fast
since primes are dense. This algorithm is in RP.



\begin{lemma}[Schwarz-Zippel]
    
\end{lemma}

\end{document}


